Annotated example of index standardization with sdmTMB

### 2021-11-15 doc/appendix-pcod-sdmTMB.Rmd

Let's perform index standardization with the built-in data for Pacific cod. The density units should be kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between v vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in (examples included in `?make_mesh`).
For example, a mesh with a cutoff distance of 10km can be made with

@@

Let's fit a GLMM. We include `0 + as.factor(year)` so that there is a factor predictor that represents the mean estimate for each time slice. Setting `silent = FALSE` would give us optimization details as the model fits but clutter the Rmd output.

```{r fit-pcod, warning=FALSE, message=FALSE}
fit <- sdmTMB(density ~ 0 + as.factor(year), data = pcod,
  time = "year", mesh = mesh, family = tweedie(link = "log"),
  silent = TRUE
)
```

Print the model:

@@

Now we want to predict on a fine-scale grid over the entire survey domain. There is a grid built into the sdmTMB package for Queen Charlotte Sound (named `qcs_grid`). Our prediction grid also needs to have all the covariates that we used in the model above.

Now make the predictions on new data.

@@

Let's make a small function to make maps.

@@

We can also look at just the fixed effects, here year:

@@

And finally we can look at the spatiotemporal random effects that represent deviation from the fixed effect predictions and the spatial random effect deviations. These represent biotic and abiotic factors that are changing through time and are not accounted for in the model.

@@

The fastest way to assess uncertainty on our spatial predictions is with simulation from the ...

@@

@@

@@

>>>>

We will fit a geostatistical spatiotemporal model with sdmTMB for the purposes of index standardization. We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations.
The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`).
Custom built meshes with `INLA` (including meshes with barriers limiting covariance across polygons) can also be passed in. See examples included in `?make_mesh`.

For example, a mesh with a cutoff distance of 10 km can be made with

@@

Next we will fit a GLMM (generalized linear mixed effects model). For this example, we will include `0 + as.factor(year)` so that there is a factor predictor that represents the mean estimate for each time slice. Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output.

```{r fit-pcod, warning=FALSE, message=FALSE}
fit <- sdmTMB(density ~ 0 + as.factor(year), data = pcod,
  time = "year", mesh = mesh, family = tweedie(link = "log")
)
```

Print the model to inspect the fit:

@@

Next, we want to predict on a fine-scale grid over the entire survey domain so we can sum up the expected biomass across the entire region each year. There is a grid built into the sdmTMB package for Queen Charlotte Sound (named `qcs_grid`). Our prediction grid also needs to have all the covariates that we used in the model above. In this case, that is a `year` column as well as `X` and `Y`.

@@

Next we will make a small function to help plot maps.

@@

We can also look at just the fixed effects, here `year`:

@@

Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations. These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model.

@@

The fastest way to assess uncertainty on our spatial predictions is with simulation from the joint precision matrix.

@@

This returns a matrix with a row for each row of the `newdata` data frame and a column for each replicate draw. We can use this matrix to calculate the standard deviation or coefficient of variation:

@@

We can calculate our standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE`. This returns the TMB objective function, which is required to pass the output to `get_index()`.

@@

The above example used the default `bias_correct = FALSE` in `get_index()`. This is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects. Setting `bias_correct = TRUE` implements the generic by his correction algorithm that should correct for this, but can take considerable time and memory.

As an alternative, we can use the output from the joint precision matrix simulation draws to calculate our index. This should approximate the output of `get_index(x, bias_correct = TRUE)`. However, we note that there may be situations where this approach may fail or be biased.
